On joint estimation of ancestral states and phylogenetic model parameters

### 2017-10-04 deprecated.tex
fix up arguments for restricted case; disjoint ideas for general case

>>>>

\subsection{Branch lengths with $x$ and $y$ unknown}

Our plan of attack is to focus on the likelihood
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
    &\qquad + p_{1}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
    &\qquad + p_{2}          \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
    &\qquad + p_{3}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
    &\qquad + p_{12}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
    &\qquad + p_{23}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
    &\qquad + p_{123}        \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
    &\qquad - \log(8) \\
    &\qquad + p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\
    &\qquad + p_{1}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
    &\qquad + p_{2}          \cdot\log((1+x')^2   (1+w')(1+y')(1-y')) \\
    &\qquad + p_{3}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
    &\qquad + p_{12}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \\
    &\qquad + p_{13}         \cdot\log((1-x')^2   (1+w')(1+y')^2) \\
    &\qquad + p_{23}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \\
    &\qquad + p_{123}        \cdot\log((1+x')^2   (1+w')(1+y')(1-y')).
\end{align*}
This simplifies to
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
    &\qquad + (p_{1} + p_{3})          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
    &\qquad + (p_{2} + p_{123})          \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
    &\qquad + (p_{12} + p_{23})         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
    &\qquad + (2-p_{1}-p_{3}-p_{12}-p_{23}-2p_{13})\log(1+x') \\
    &\qquad + (p_{1}+p_{3}+p_{12}+p_{23}+2p_{13})\log(1-x') \\
    &\qquad + (2-p_{2}-p_{12}-p_{23}-p_{123})\log(1+y') \\
    &\qquad + (p_{2}+p_{12}+p_{23}+p_{123})\log(1-y') \\
    &\qquad + \log(1+w').
\end{align*}
This likelihood occurs when $\hat{w}' = 1$ and $\hat{y}' > \hat{x}'$, the latter coming from
$$
(1-x'^2)(1+y'^2) > (1+x'^2)(1-y'^2)
$$
only if $y' > x'$.
We simply need to obtain
$$
(\hat{x}', \hat{y}', \hat{w}') = \arg\max_{x, y, w} \ \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
$$
as functions of $x$ and $y$ and verify the region where $\hat{w}'(x, y) = 1$ and $\hat{y}'(x, y) > \hat{x}'(x, y)$.
This region will be consistent with one always estimating $\hat{w}'$ to be one.
This can similarly be done for the other likelihood term with $\hat{x}'(x, y) > \hat{y}'(x, y)$.

To do is to find these functions---or subsets of their varieties---in as closed a form as possible.

When we are in the region where $\hat{w}'$ is one, necessarily (I think) our region is going to be a subset of the case where $x'$ and $y'$ are fixed.

### 2017-10-04 joint_inf.tex
fix up arguments for restricted case; disjoint ideas for general case

\begin{figure}
\centering
\includegraphics[width=.9\textwidth]{analytic-inconsistency}
\caption{
    Regions of inconsistency.
    Due to the looseness of the upper and lower bounds, the parameters in the top right do not necessarily indicate consistency, though all parameters in the labeled region result in an inconsistency.
}
\label{fig:inconsistency-farris}
\end{figure}

@@

First, assume we know the true values of $x'$ and $y'$ but not $w'$.
In other words, we fix $x'=x$ and $y'=y$ to their true, generating values and wish to estimate the internal branch parameter $w'$.
In this case, the likelihood as a function of $w'$ is proportional to either
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
    &\qquad + \log((1+w'))
\end{align*}
or
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
    &\qquad + (1 - p_{13})\cdot\log((1+w')) \\
    &\qquad + p_{13}\cdot\log((1-w')).
\end{align*}
When computing the global minimum for $0 \le w' \le 1$ using basin-hopping, we obtain the fit $\hat{w}$ in Fig.~\ref{fig:w_hat}.
We find that the estimate for $\hat{w}$ will typically be an overestimate, and in fact is almost always identically equal to 1, indicating no change along that particular branch.

\begin{figure}
\centering
    \begin{tabular}{cc}
\includegraphics[width=.45\textwidth]{w_hat_heatmap} &
\includegraphics[width=.45\textwidth]{w_hat_heatmap_no_joint}\\
Joint inference & Marginal inference \\
    \end{tabular}
\caption{
    Estimating $\hat{w}$ for $x'$ and $y'$ fixed at their true values.
}
\label{fig:w_hat}
\end{figure}

Second, we consider the more standard case where the true values of $x$ and $y$ are not known.
The form of the likelihood will alter the fitted values of $x$ and $y$ so that even with perfect information we will not necessarily estimate $\hat{x}=x$ and $\hat{y}=y$.
If we estimate $(x', y', w')$ jointly using the same approach above, we obtain the fit $\hat{w}$ in Figure~\ref{fig:w_hat_joint}.

\begin{figure}
\centering
\includegraphics[width=.9\textwidth]{w_hat_heatmap_joint}
\caption{
    Estimating $(\hat{x}, \hat{y}, \hat{w})$.
}
\label{fig:w_hat_joint}
\end{figure}

We no longer consistently overestimate $\hat{w}$, but the estimates are surprisingly unstable, with estimates of 1 for $\hat{w}$ scattered throughout the $(x,y)$ lattice.
In summary, even given the correct Farris topology we obtain poor estimates for the value of the interior branch length.

>>>>

\begin{figure}
\centering
\includegraphics[width=.9\textwidth]{topology-inconsistency-legend}
\caption{
    Regions of inconsistency.
    Due to the looseness of the upper and lower bounds, the parameters in the top right do not necessarily indicate consistency, though all parameters in the labeled region result in an inconsistency.
}
\label{fig:inconsistency-farris}
\end{figure}

@@

First, assume we know the true values of $x'$ and $y'$ but not $w'$.
In other words, we fix $x'=x$ and $y'=y$ to their true, generating values and wish to estimate the internal branch parameter $w'$.
In this case, the likelihood as a function of $w'$ is proportional to either
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
    &\qquad + \log(1+w')
\end{align*}
or
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
    &\qquad + (1 - p_{13})\cdot\log(1+w') \\
    &\qquad + p_{13}\cdot\log(1-w').
\end{align*}
Since
$$
\log(1+w') >= (1 - p_{13})\cdot\log(1+w') + p_{13}\cdot\log(1-w')
$$
with equality only when $w'=0$, we focus on the first likelihood.
Simplifying and suppressing some variables, this likelihood can be written
$$
\ell(w') = \alpha_1\log(\beta+\gamma w') + \alpha_2\log(\beta-\gamma w') + \log(1+w')
$$
where
$$
\alpha_1 := \alpha_1(x, y) = 1+x^2+y^2+4xy^2+x^2y^2,
$$
$$
\alpha_2 := \alpha_2(x, y) = 1+x^2+y^2-4xy^2+x^2y^2,
$$
$$
\beta := \beta(x, y) = 1+x^2+y^2+x^2y^2,
$$
$$
\gamma := \gamma(x, y) = 4xy.
$$
Since $x$ and $y$ fall between zero and one, we make use of the inequalities (excepting the cases of $x=y=0$ and $x=y=1$)
$$
\alpha_1 > \alpha_2
$$
and
$$
\beta > \gamma.
$$
% Not sure if I need to show this, but if x and y aren't 1 it's 1+x^2+y^2+x^2y^2 > 4xy => (1+x^2)(1+y^2) > 4xy and 1+x^2 > 2x, 1+y^2 > 2y qed
The derivative is
$$
\ell'(w') := \frac{d}{dw'} \ell(w') = \frac{\alpha_1 \gamma}{\beta+\gamma w'} + \frac{\alpha_2 \gamma}{\beta-\gamma w'} + \frac{1}{1+w'}.
$$
The inequality $\beta > \gamma$ implies that any inequality relating $\ell'$ and 0 is equivalent to the same inequality relating 0 and the quadratic function
\begin{align*}
f(w') &= w^2\cdot(-\gamma^2\alpha_1-\gamma^2\alpha_2-\gamma^2) \\
      &\qquad + w\cdot(\gamma\alpha_1\beta-\gamma^2\alpha_1-\gamma\alpha_2\beta-\gamma^2\alpha_2) \\
      &\qquad + (\gamma\alpha_1\beta-\gamma\alpha_2\beta+\beta^2), \\
\end{align*}
implying $\ell'$ has two zeros according to the quadratic formula.
By $\alpha_1 > \alpha_2$, $\ell'(0) > 0$ implies $\ell$ is increasing at $w'=0$.
We use this fact to show that, if the smaller of the zeros of $f(w')$ is greater than one, $\hat{w}' \equiv 1$.
Using the quadratic formula with
$$
a(x, y) = -\gamma^2 \alpha_1 - \gamma^2 \alpha_2 - \gamma^2,
$$
$$
b(x, y) = \gamma \alpha_1 \beta - \gamma^2\alpha_1 - \gamma \alpha_2 \beta - \gamma^2 \alpha_2,
$$
$$
c(x, y) = \gamma \alpha_1 \beta - \gamma \alpha_2 \beta + \beta^2,
$$
the smaller zero is
$$
\hat{w}'(x, y) = \frac{-b(x, y) - \sqrt{b(x, y)^2 - 4a(x, y)c(x, y)}}{2a(x, y)},
$$
which is a function of the generating parameters $x$ and $y$.
We plot the region where $\hat{w}'(x, y) \ge 1$ in Figure~\ref{fig:bl-inconsistency}.

\begin{figure}
\centering
\includegraphics[width=.9\textwidth]{branch-length-inconsistency-legend}
\caption{Region of inconsistency with $x$ and $y$ fixed}
\label{fig:bl-inconsistency}
\end{figure}

We consider the more standard case where the true values of $x$ and $y$ are not known.
The form of the likelihood will alter the fitted values of $x$ and $y$ so that even with perfect information we will not necessarily estimate $\hat{x}'=x$ and $\hat{y}'=y$.
When computing the global maximum for $0 \le x',y',w' \le 1$ using basin-hopping, we obtain the fit $\hat{w}'$ in Fig.~\ref{fig:w_hat_joint}.
The region in which $\hat{w}'$ is estimated to be one---indicating no change along that particular branch---is now much smaller, but still nontrivial.
Additionally, the estimates are surprisingly unstable, with estimates of 1 for $\hat{w}'$ scattered throughout the $(x,y)$ lattice.
Even given the correct Farris topology we have a region of the space of true parameters where we will always estimate the interior branch parameter to be one.

\begin{figure}
\centering
\includegraphics[width=.9\textwidth]{w_hat_heatmap_joint}
\caption{
    Estimating $(\hat{x}', \hat{y}', \hat{w}')$.
}
\label{fig:w_hat_joint}
\end{figure}
